---
title: "Compare Core vs. Soft-core vs. Non-core distances (ST1 clade2_R20291)"
author: "Skylar Zeng"
format: html
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(readr)
library(here)
library(ape)
library(patchwork)
```

## Import data
```{r, echo=FALSE}
# FASTA alignments
aln_var <- read.dna(
  here("variant_calling", "clade2_R20291", "phylo_analysis", "gubbins", "clade2_R20291_gubbins_masked_var_sites.fa"),
  format = "fasta"
)

# Optional list of "secondary isolates" to drop later (pairs de-dup)
secondary_isolates <- read_tsv(
  here("data", "metadata", "secondary_isolate_ids.txt"),
  col_names = FALSE
)$X1
```

## Helper functions
```{r, include=FALSE}
# Force any object to a 2D matrix
.ensure_matrix <- function(x, nrow = NULL, ncol = NULL) {
  if (!is.null(dim(x))) return(x)
  if (!is.null(nrow) && !is.null(ncol)) return(matrix(x, nrow = nrow, ncol = ncol))
  # default: make it a 1-row matrix
  matrix(x, nrow = 1)
}

# DNAbin -> character matrix (samples x sites), works for 1 seq or 1 site too
to_char_mat <- function(dna) {
  x <- as.character(dna)             # ape may return list, matrix, or vector
  if (is.matrix(x)) {
    return(x)
  } else if (is.list(x)) {
    m <- do.call(rbind, x)           # rows = samples, cols = sites
    return(.ensure_matrix(m))
  } else if (is.vector(x)) {
    m <- matrix(x, nrow = 1)         # 1 sample, length(x) sites
    return(m)
  } else {
    stop("Unexpected as.character(DNAbin) output.")
  }
}

# Soft-core filter: keep columns with <= threshold missing across samples
soft_core_filter <- function(dna, threshold = 0.05) {
  char_mat  <- to_char_mat(dna)                              # samples x sites
  char_mat  <- .ensure_matrix(char_mat)
  # Treat N/n/-/?/. as missing
  miss      <- char_mat %in% c("N","n","-","?","." )
  miss      <- .ensure_matrix(miss, nrow = nrow(char_mat), ncol = ncol(char_mat))

  # If there are zero columns, bail gracefully
  if (ncol(miss) == 0) {
    warning("Alignment has 0 columns; nothing to keep.")
    return(list(aln = dna[, integer(0), drop = FALSE],
                keep = logical(0), n_kept = 0, n_all = 0, prop = numeric(0)))
  }

  prop_miss <- colMeans(miss)                                 # length = ncol
  keep      <- prop_miss <= threshold

  # If nothing passes, still return a valid empty-alignment result
  if (!any(keep)) {
    warning("Soft-core filter kept 0 columns; check threshold or alignment.")
    return(list(aln = dna[, integer(0), drop = FALSE],
                keep = keep, n_kept = 0, n_all = ncol(char_mat), prop = prop_miss))
  }

  list(
    aln    = dna[, keep, drop = FALSE],
    keep   = keep,
    n_kept = sum(keep),
    n_all  = ncol(char_mat),
    prop   = prop_miss
  )
}

# Distance -> matrix; drop the reference if present
convert_dist_to_matrix <- function(dist_obj, ref_name = "cdiff_R20291_ref_genome") {
  mat <- as.matrix(dist_obj)
  if (!is.null(mat) && length(mat)) {
    if (ref_name %in% rownames(mat)) {
      mat <- mat[rownames(mat) != ref_name, colnames(mat) != ref_name, drop = FALSE]
    }
  }
  mat
}

# Symmetric matrix to canonical long (id1<=id2), unique pairs, drop self
mat_to_long_pairs <- function(m, drop_ref = TRUE, ref_name = "cdiff_R20291_ref_genome") {
  # handle empty/NULL matrices gracefully
  if (is.null(m) || is.null(dim(m)) || any(dim(m) == 0)) {
    return(tibble(id1 = character(), id2 = character(), snvs = double()))
  }
  # optionally drop reference
  if (drop_ref && !is.null(rownames(m)) && ref_name %in% rownames(m)) {
    m <- m[rownames(m) != ref_name, colnames(m) != ref_name, drop = FALSE]
    if (any(dim(m) == 0)) {
      return(tibble(id1 = character(), id2 = character(), snvs = double()))
    }
  }
  rn <- rownames(m); cn <- colnames(m)
  if (is.null(rn) || is.null(cn)) stop("Matrix needs row and column names.")

  # build all pairs, attach values
  g <- expand.grid(id1 = rn, id2 = cn, stringsAsFactors = FALSE)
  g$snvs <- m[cbind(match(g$id1, rn), match(g$id2, cn))]

  # canonicalize to undirected pairs, drop self, de-dup
  g %>%
    filter(id1 != id2) %>%
    mutate(a = pmin(id1, id2), b = pmax(id1, id2)) %>%
    transmute(id1 = a, id2 = b, snvs = snvs) %>%
    distinct(id1, id2, .keep_all = TRUE)
}
```

# 5% soft-core
## Distances: core (strict), non-core (pairwise), soft-core (≤5% missing)
```{r, echo=FALSE}
# Core: strict use of all columns in aln_var with pairwise.deletion = FALSE
var_dist_core     <- dist.dna(aln_var, pairwise.deletion = FALSE, model = "N")

# Non-core (pairwise-inclusive): same alignment, pairwise.deletion = TRUE
var_dist_noncore  <- dist.dna(aln_var, pairwise.deletion = TRUE,  model = "N")

# Soft-core: filter columns by missingness threshold, then pairwise.deletion = TRUE
aln_var_soft      <- soft_core_filter(aln_var, threshold = 0.05)
cat("Soft-core retained: ", aln_var_soft$n_kept, " / ", aln_var_soft$n_all, " columns\n", sep = "")
var_dist_softcore <- dist.dna(aln_var_soft$aln, pairwise.deletion = TRUE, model = "N")
```

## Matrices & exports
```{r, echo=FALSE}
# Convert to matrices and drop reference if present
var_core_matrix     <- convert_dist_to_matrix(var_dist_core)
var_softcore_matrix <- convert_dist_to_matrix(var_dist_softcore)
var_noncore_matrix  <- convert_dist_to_matrix(var_dist_noncore)

# Save matrices
dir.create(here("data","SNV_matrices"), showWarnings = FALSE, recursive = TRUE)

write.table(var_core_matrix,
            file = here("data","SNV_matrices","clade2_r20291_core_dist_matrix.tsv"),
            sep = "\t", quote = FALSE, col.names = NA)
write.table(var_softcore_matrix,
            file = here("data","SNV_matrices","clade2_r20291_softcore_dist_matrix.tsv"),
            sep = "\t", quote = FALSE, col.names = NA)
write.table(var_noncore_matrix,
            file = here("data","SNV_matrices","clade2_r20291_noncore_dist_matrix.tsv"),
            sep = "\t", quote = FALSE, col.names = NA)
```

## Long-form (pairs), clean, and save
```{r, echo=FALSE}
core_long     <- mat_to_long_pairs(var_core_matrix)
softcore_long <- mat_to_long_pairs(var_softcore_matrix)
noncore_long  <- mat_to_long_pairs(var_noncore_matrix)

# Drop secondary isolates if provided
core_clean <- core_long %>%
  filter(!(id1 %in% secondary_isolates), !(id2 %in% secondary_isolates))
softcore_clean <- softcore_long %>%
  filter(!(id1 %in% secondary_isolates), !(id2 %in% secondary_isolates))
noncore_clean <- noncore_long %>%
  filter(!(id1 %in% secondary_isolates), !(id2 %in% secondary_isolates))

write_csv(core_clean,     here("data","SNV_matrices","clade2_r20291_core_dist_long.csv"))
write_csv(softcore_clean, here("data","SNV_matrices","clade2_r20291_softcore_dist_long.csv"))
write_csv(noncore_clean,  here("data","SNV_matrices","clade2_r20291_noncore_dist_long.csv"))
```

## Compare pairwise distances (core vs soft-core vs non-core)
```{r, echo=FALSE}
core_for_merge     <- dplyr::rename(core_clean,     snvs_core     = snvs)
softcore_for_merge <- dplyr::rename(softcore_clean, snvs_softcore = snvs)
noncore_for_merge  <- dplyr::rename(noncore_clean,  snvs_noncore  = snvs)

snv_comparison <- core_for_merge %>%
  inner_join(softcore_for_merge, by = c("id1","id2")) %>%
  inner_join(noncore_for_merge,  by = c("id1","id2")) %>%
  mutate(
    delta_soft_vs_core    = snvs_softcore - snvs_core,
    delta_noncore_vs_soft = snvs_noncore  - snvs_softcore
  )

# Summaries
cat("Pairs changed (Core → Soft-core): ",
    sum(snv_comparison$delta_soft_vs_core != 0), "\n",
    "Max Δ (Soft - Core): ",
    ifelse(nrow(snv_comparison)>0, max(snv_comparison$delta_soft_vs_core), NA), "\n",
    "Mean Δ (Soft - Core): ",
    ifelse(nrow(snv_comparison)>0, sprintf('%.3f', mean(snv_comparison$delta_soft_vs_core)), NA), "\n\n",
    "Pairs changed (Soft-core → Non-core): ",
    sum(snv_comparison$delta_noncore_vs_soft != 0), "\n",
    "Max Δ (Non-core - Soft): ",
    ifelse(nrow(snv_comparison)>0, max(snv_comparison$delta_noncore_vs_soft), NA), "\n",
    "Mean Δ (Non-core - Soft): ",
    ifelse(nrow(snv_comparison)>0, sprintf('%.3f', mean(snv_comparison$delta_noncore_vs_soft)), NA), "\n",
    sep = ""
)
```
### Core vs Soft-core — scatter
```{r, echo=FALSE}
ggplot(snv_comparison, aes(x = snvs_core, y = snvs_softcore)) +
  geom_point(size = 2) +
  coord_cartesian(xlim = c(0, 100)) +
  labs(
    x = "Core SNVs",
    y = "Soft-core SNVs",
    title = "SNV Distance: Core vs Soft-core (Clade 2)"
  ) +
  theme_minimal()
```
### Soft-core vs Non-core — scatter
```{r, echo=FALSE}
ggplot(snv_comparison, aes(x = snvs_softcore, y = snvs_noncore)) +
  geom_point(size = 2) +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 300)) +
  labs(
    x = "Soft-core SNVs",
    y = "Non-core (pairwise) SNVs",
    title = "SNV Distance: Soft-core vs Non-core (0–100) (Clade 2)"
  ) +
  theme_minimal()
```
## Bonus: delta histograms (how much distances change)
```{r, echo=FALSE}
# Core → Soft-core
ggplot(snv_comparison, aes(x = delta_soft_vs_core)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Δ (Soft-core - Core)", y = "Pairs", title = "Distance Change: Core → Soft-core") +
  theme_minimal()

# Soft-core → Non-core
ggplot(snv_comparison, aes(x = delta_noncore_vs_soft)) +
  geom_histogram(binwidth = 1) +
  labs(x = "Δ (Non-core - Soft-core)", y = "Pairs", title = "Distance Change: Soft-core → Non-core") +
  theme_minimal()
```
```{r, echo=FALSE}
library(dplyr)
library(tidyr)

# helper: non-overlapping bin counts
count_bins <- function(x) {
  tibble(
    `0-2`   = sum(x >= 0  & x <= 2,  na.rm = TRUE),
    `2-5`   = sum(x >  2  & x <= 5,  na.rm = TRUE),
    `5-10`  = sum(x >  5  & x <= 10, na.rm = TRUE),
    `10-20` = sum(x > 10  & x <= 20, na.rm = TRUE)
  )
}

summary_counts <- bind_rows(
  count_bins(snv_comparison$snvs_core)     %>% mutate(set = "core"),
  count_bins(snv_comparison$snvs_softcore) %>% mutate(set = "soft-core"),
  count_bins(snv_comparison$snvs_noncore)  %>% mutate(set = "non-core")
) %>%
  select(set, `0-2`, `2-5`, `5-10`, `10-20`) %>%
  mutate(set = factor(set, levels = c("core","soft-core","non-core"))) %>%
  arrange(set)

summary_counts
```

# Soft-core thredhold vs. Proportion of positions
```{r, echo=FALSE}
library(tidyverse)

# DNAbin -> character matrix (samples x sites), robust to 1-sample / 1-site
to_char_mat <- function(dna) {
  x <- as.character(dna)                     # list / matrix / vector depending on ape
  if (is.matrix(x)) {
    m <- x
  } else if (is.list(x)) {
    m <- do.call(rbind, x)
  } else if (is.vector(x)) {
    m <- matrix(x, nrow = 1)
  } else {
    stop("Unexpected DNAbin -> character form")
  }
  # Guarantee 2D
  if (is.null(dim(m))) m <- matrix(m, nrow = 1)
  m
}

# --- Build per-site missingness (from aln_var) ---
var_char <- to_char_mat(aln_var)                  # samples x sites (characters)
miss_mat <- var_char %in% c("N","n","-","?",".")

# Force dimensions in case R drops them (e.g., 1-column)
if (is.null(dim(miss_mat))) {
  miss_mat <- matrix(miss_mat, nrow = nrow(var_char), ncol = ncol(var_char))
}

total_sites <- if (!is.null(dim(var_char))) ncol(var_char) else length(var_char)
if (total_sites == 0) {
  stop("aln_var has 0 columns (no sites). Check your input FASTA.")
}

prop_miss <- colMeans(miss_mat)                   # length = #sites

# --- Sweep thresholds and compute % included ---
thresholds <- seq(0, 0.5, by = 0.01)              # 0% to 50% missing allowed
th_summary <- tibble(
  threshold = thresholds,
  n_included = vapply(thresholds, function(t) sum(prop_miss <= t), integer(1))
) |>
  mutate(pct_included = 100 * n_included / total_sites)

# --- Compact table at key thresholds ---
key_th <- c(0, 0.01, 0.02, 0.05, 0.10, 0.20, 0.30)
th_table <- th_summary |>
  filter(threshold %in% key_th) |>
  transmute(
    `Soft-core threshold (≤ missing)` = paste0(round(threshold*100, 2), "%"),
    `Positions included` = n_included,
    `Included (%)` = round(pct_included, 2)
  )

th_table
```

```{r, echo=FALSE}
five_pct <- th_summary |> filter(threshold == 0.05)
cat("Soft-core 5% → included:", five_pct$n_included, "of", total_sites,
    sprintf("(%.2f%%)\n", five_pct$pct_included))

## --- Find threshold that includes at least 70% of positions ---
target <- 0.70
sorted <- sort(prop_miss, na.last = NA)
idx    <- max(1, ceiling(target * length(sorted)))
t70    <- sorted[idx]                           # missingness threshold (fraction) to reach ≥70% included
inc70  <- sum(prop_miss <= t70)
pct70  <- 100 * inc70 / total_sites

# One-row summary table
tibble(
  `Target included`          = "70%",
  `Threshold (≤ missing)`    = sprintf("%.2f%%", t70 * 100),
  `Positions included`       = inc70,
  `Included (%)`             = round(pct70, 2)
)
```

```{r, echo=FALSE}
ggplot(th_summary, aes(x = threshold*100, y = pct_included)) +
  geom_line() +
  geom_point(data = dplyr::filter(th_summary, threshold %in% key_th)) +
  geom_vline(xintercept = 5, linetype = "dashed") +              # your usual soft-core (5%)
  geom_vline(xintercept = t70*100, linetype = "dotted") +        # threshold to reach 70% coverage
  geom_point(aes(x = t70*100, y = pct70), inherit.aes = FALSE) +
  annotate("text",
           x = t70*100, y = pct70,
           label = sprintf("70%% at %.2f%%", t70*100),
           vjust = -0.6, size = 3) +
  labs(
    x = "Soft-core missingness threshold (%)",
    y = "Positions included (%)",
    title = "Coverage vs. soft-core threshold (Clade 2)"
  ) +
  theme_minimal()
```

