---
title: "Initial exploration on patient location data"
author: "Skylar Zeng"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("readr")
library("tidyverse")
library("readxl")
library("here")
library("RColorBrewer")
library("knitr")
library("ggthemes")
```

# Purpose
The purpose of this notebook is to tabulate and visualize patient and location data of interest for the UM Sys Bio C diff data. They are: 

  1. Descriptive statistics: patient data (age, gender, race, LOS)
  2. hospital vs. community onset of infection by MLST
  3. for inpatient infections, days between admission and infection onset by MLST
  4. Locations variation and temporal overlap

```{r import-data, warning = FALSE, message = FALSE}
############################
# import relevant datasets
############################
ids <- read_csv(here('data','metadata','cleaned_passed_genomeid.csv'), col_names = FALSE)$X1

patient_data <- read_csv(here("data","metadata", "hospital_onset_data.csv")) %>%
  filter(genome_id %in% ids)

patient_data_tx <- read_excel(here("data","metadata", "summary_toxin_metadata.xlsx"))

mlst_data <- read_tsv(here("data", "metadata", "mlst_replaced.tsv")) %>%
  mutate(ST = str_remove(ST, "\\*"))

main_wiens<- read_tsv(here("data", "metadata", "main_from_wiens_group_filtered_cdiff_cases.tsv")) %>%
  arrange(PatientID, Date)

# patient locations
locations <- read_tsv(here('data', 'metadata', 'locations_from_wiens_group_filtered.tsv'))
```

```{r}
#subset of patients that we want demographics on
main_wiens_subset <- main_wiens %>% 
  filter(PatientID %in% patient_data_tx$RDW_patient_id) # combine patient and wiens data

# Convert to Date class to keep only year-month-day (drop time)
main_wiens_subset$CollectionDateCdiff <- as.Date(main_wiens_subset$CollectionDateCdiff)
patient_data_tx$collection_date <- as.Date(patient_data_tx$collection_date)

# Perform the full join on PatientID and collection date
combined_data <- inner_join(
  main_wiens_subset,
  patient_data_tx,
  by = c("PatientID" = "RDW_patient_id", "CollectionDateCdiff" = "collection_date")
)

# keep only one isolate
combined_data_single_isolate <- combined_data %>% 
  filter(isolate == "a") %>%
  filter(sample_type %in% c("case_formed", "case_unformed"))
```

```{r merge-datasets, warning = FALSE, message = FALSE, echo = FALSE}
##############################
# to those that are sequenced
##############################
patient_and_mlst <- inner_join(
  patient_data_tx,
  mlst_data,
  by = c("genomeID" = "genome_id")
)
```

# Demographics (n=727)
## Age distribution (at least 18 yrs-old)
**Interpretation:** Strong skew toward older adults. This pattern is consistent with the known epidemiology of C. difficile, where older individuals — particularly those over 65 — are at increased risk due to more frequent healthcare exposure, antibiotic use, and comorbidities.
```{r}
summary(combined_data_single_isolate$AdmitAge)
```

```{r plot-age-distribution, warning = FALSE, message = FALSE, echo = FALSE}
ggplot(combined_data_single_isolate,
       aes(x = AdmitAge)) + 
       geom_histogram(stat = "count")
```

## Gender
```{r}
table(combined_data_single_isolate$GenderCode)
```

## Race
**Interpretation:** The dataset is overwhelmingly composed of Caucasian patients, who represent 84.0% (604/719) of the total. African American patients make up the second-largest group at 10.8% (78). All other racial groups are represented in small numbers
```{r}
table(combined_data_single_isolate$RaceName)
```
```{r}
# Convert and calculate percent
race_df <- as.data.frame(table(combined_data_single_isolate$RaceName))
colnames(race_df) <- c("Race", "Count")
race_df <- race_df %>%
  mutate(Percent = round(100 * Count / sum(Count), 1),
         Label = paste0(Count, " (", Percent, "%)"))

# Plot with labels
ggplot(race_df, aes(x = reorder(Race, -Count), y = Count, fill = Race)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  geom_text(aes(label = Label), vjust = -0.5, size = 3.5) +
  theme_minimal() +
  labs(title = "Race Distribution of Patients",
       x = "Race/Ethnicity",
       y = "Number of Patients") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## Length of stay (days)
**Interpretation:** 
- The peak in frequency occurs within the first 5 days, indicating that short stays are most common.
- Median LOS is 8 days, with an interquartile range (IQR) from 5 to 18 days. 
- Several patients had very prolonged hospitalizations, including stays over 100 days, and a maximum LOS of 350 days.
```{r}
summary(combined_data_single_isolate$LOS)
ggplot(combined_data_single_isolate,
       aes(x = LOS)) + 
       geom_histogram(stat = "count")
```

## Infection Onset Type (0 = Community-Onset, 1 = Hospital-Onset)
```{r}
# Check raw counts (optional)
table(combined_data_single_isolate$hospital_onset)

hospital_onset_summary <- combined_data_single_isolate %>%
  filter(!is.na(hospital_onset)) %>%
  mutate(
    hospital_onset = ifelse(hospital_onset == 1, "Hospital-Onset", "Community-Onset"),
    hospital_onset = factor(hospital_onset, levels = c("Community-Onset", "Hospital-Onset"))
  ) %>%
  count(hospital_onset, name = "n") %>%
  mutate(
    p = n / sum(n),
    perc = scales::percent(p, accuracy = 0.1),
    label = paste0("n=", n, " (", perc, ")")
  )

ggplot(hospital_onset_summary, aes(x = hospital_onset, y = n, fill = hospital_onset)) + 
  geom_col() +
  geom_text(aes(label = label), vjust = -0.5) +
  scale_fill_manual(values = c("Community-Onset" = "#F0E442", "Hospital-Onset" = "#D55E00")) +
  labs(x = "Onset Type", y = "Count", fill = "Onset",
       title = "Distribution of Hospital vs. Community Onset Cases") +
  theme_minimal() +
  expand_limits(y = max(hospital_onset_summary$n) * 1.15)  # avoid label clipping
```

## Inpatients vs. Outpatient (1 == inpatient)
```{r plot-inpatients, echo = FALSE, message = FALSE, warning = FALSE}

# Filter out NA and calculate counts and percentages
inpatient_summary <- combined_data_single_isolate %>%
  filter(!is.na(inpatient)) %>%
  mutate(inpatient = recode(as.factor(inpatient), `0` = "Outpatient", `1` = "Inpatient")) %>%
  count(inpatient) %>%
  mutate(perc = scales::percent(n / sum(n), accuracy = 0.1),
         label = paste0("n=", n, " (", perc, ")"))

# Plot
ggplot(inpatient_summary, aes(x = inpatient, y = n, fill = inpatient)) + 
  geom_col() +
  geom_text(aes(label = label), vjust = -0.5) +
  scale_fill_manual(values = c("Outpatient" = "#E69F00", "Inpatient" = "#56B4E9")) +
  labs(x = "Patient Status", y = "Count", fill = "Status",
       title = "Distribution of Isolates by Inpatient Status") +
  theme_minimal()
```
## ST type
```{r plot-mlst-types, fig.width=15, message = FALSE, warning = FALSE}
patient_and_mlst <- inner_join(patient_data, mlst_data, by = "genome_id")

patient_and_mlst$sorted_ST <- factor(patient_and_mlst$ST, 
                              levels = names(sort(table(patient_and_mlst$ST), 
                              decreasing=FALSE)))

only_one <- names(which(table(patient_and_mlst$ST) == 1))

patient_and_mlst_filtered <- filter(patient_and_mlst,
                                    !(is.na(sorted_ST)) & !(sorted_ST %in% only_one))
ggplot(patient_and_mlst,
       aes(x = sorted_ST)) + 
       geom_histogram(stat = "count")
# Prepare color vector
n_STs <- length(unique(patient_and_mlst_filtered$sorted_ST))

# Choose a palette (e.g., "Paired", "Set3", "Dark2")
library(RColorBrewer)
ST_colors <- colorRampPalette(ggthemes_data[["tableau"]][["color-palettes"]][["regular"]]$`Tableau 20`$value)(69)

# Plot with different colors for each ST
ST_bar_plot <- ggplot(patient_and_mlst_filtered, 
       aes(x = sorted_ST, fill = sorted_ST)) +
       geom_bar(stat = "count") +
       scale_fill_manual(values = ST_colors) +  # Apply BMJ categorical fill palette
       coord_flip() +
       theme_light() +
       theme(axis.text.x = element_text(angle = 90),
             text = element_text(size = 22)) +
       ggtitle("Most common sequence types in comprehensive cohort") +
       ylab("Isolate count") +
       xlab("Sequence type")

ST_bar_plot 
```

```{r generate-recurrence-variable}
patient_and_mlst <- patient_and_mlst %>%
                      group_by(subject_id) %>%
                      mutate(secondary_isolate = ifelse(collection_date == min(collection_date), 0, 1) )

table(patient_and_mlst$secondary_isolate)

# Export list of secondary isolates (more than 1 isolate per person, not taking into account timing)
write(patient_and_mlst$genome_id[patient_and_mlst$secondary_isolate == 1], here("data", "metadata", "secondary_isolate_ids.txt"))

# Import secondary isolates
secondary_isolates <- read_tsv(here("data", "metadata", "secondary_isolate_ids.txt"), col_names = FALSE)$X1
```

Repeat above plot with secondary isolates highlighted

```{r plot-mlst-types-sorted, fig.width=15, message = FALSE, warning = FALSE}
patient_and_mlst$sorted_ST <- factor(patient_and_mlst$ST, 
                              levels = names(sort(table(patient_and_mlst$ST), 
                              decreasing=FALSE)))

only_one <- names(which(table(patient_and_mlst$ST) == 1))

patient_and_mlst$secondary <- factor(ifelse((patient_and_mlst$genome_id %in% secondary_isolates), "secondary isolate","first isolate"), levels = c(c("secondary isolate", "first isolate")))


patient_and_mlst_filtered <- filter(patient_and_mlst,
                                    ! (is.na(sorted_ST)) & !(sorted_ST %in% only_one))

ST_bar_plot_repeats <- ggplot(patient_and_mlst_filtered, 
       aes(x = sorted_ST, fill = secondary)) + 
       geom_bar(stat = "count") +
       theme(axis.text.x = element_text(angle = 90)) + 
       ggtitle("Most common sequence types in comprehensive cohort (N = 1538)") +
       coord_flip() + 
       theme_light() + 
       ylab("Isolate count") +
       xlab("Sequence type") +
       theme(text = element_text(size = 22),
             legend.title = element_blank(),
             legend.position = c(0.5,0.5)) + 
       scale_fill_manual(values = c("#DF7a26","#316593"))

ST_bar_plot_repeats
```

## ST on HO vs. CO

```{r}
# Merge ST in with patient data
patient_data_unique_ST_patient <- left_join(patient_data, mlst_data, by = c("genome_id")) %>%
  distinct(RDW_patient_id, ST, .keep_all= TRUE)

patient_data_unique_ST_patient <- patient_data_unique_ST_patient %>%
  filter(!is.na(ST),
         RDW_patient_id %in% main_wiens$PatientID) %>%
  group_by(ST) %>%
  mutate(binned_ST = ifelse(n() > 1, ST, "Singletons"))

patient_data_unique_ST_patient$sorted_ST <- factor(patient_data_unique_ST_patient$binned_ST, 
                                                   levels = names(sort(table(patient_data_unique_ST_patient$binned_ST), 
                                                                       decreasing=FALSE))) # make sorted ST for plotting

patient_data_unique_ST_patient$sorted_ST <- fct_relevel(patient_data_unique_ST_patient$sorted_ST, "Singletons", after = 0)

# make color pallette
colors_all <- colorRampPalette(ggthemes_data[["tableau"]][["color-palettes"]][["regular"]]$`Tableau 20`$value)(78)

# plot 
figure1b <- ggplot(patient_data_unique_ST_patient) +
  geom_bar(
    aes(x = sorted_ST, fill = factor(hospital_onset, levels = c(0, 1))),
    stat = "count",
    alpha = 0.9
  ) +
  labs(x = "ST") + 
  coord_flip() + 
  theme_light() + 
  ylab("Isolate count") +
  xlab("Sequence type") + 
  scale_y_continuous(breaks = seq(0, 110, 10)) + 
  theme(
    plot.margin = unit(c(1.5, 1.5, 1.5, 1.5), "cm"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    axis.text.x = element_text(size = 14)
  ) +
  scale_fill_manual(
    name = "Infection onset type",
    values = c("#77AADD", "#114477"),
    labels = c("Community-onset (CO)", "Healthcare-onset (HO)")
  )

plot(figure1b)
```

For the rest of the plots/tables, I will only keep specific bins for ST types for which we have more than 50 isolates. The rest will get grouped in an "other" bin, along with the novel and non-determined sequence types according to Ariba. 

```{r group-sequence-types, echo = FALSE}
##########################################
# Create a grouped sequence type variable 
# Keep ST types with >50 isolates, 
# group others (incl. "NA") as "other"
##########################################

# Count isolates per ST and filter those with >50
# Count isolates per ST, excluding NAs
st_counts <- table(na.omit(patient_and_mlst$ST))
over_50_isolates <- names(st_counts[st_counts > 50])

# Create grouped ST variable
patient_and_mlst <- patient_and_mlst %>%
  mutate(grouped_st = ifelse(ST %in% over_50_isolates, ST, "other"))
```

Next I took the most common sequence types and broke down those counts by inpatient vs. not (again inpatient == 1)

```{r plot-st-by-inpatient, message = FALSE, warning = FALSE, fig.height = 6}
ST_bar_plot_inpatients_all_st <- ggplot(filter(patient_and_mlst_filtered, !is.na(inpatient)), 
       aes(x = sorted_ST, fill = factor(inpatient, levels = c(0,1)))) + 
       geom_bar(stat = "count") +
       theme(axis.text.x = element_text(angle = 90)) + 
       ggtitle("STs collected by inpatient/outpatient") +
       coord_flip() + 
       theme_light() + 
       ylab("Isolate count") +
       xlab("Sequence type") +
       theme(text = element_text(size = 22),
             legend.title = element_blank(),
             legend.position = c(0.5,0.5),
             legend.key.size = unit(20, "pt")) + 
       scale_fill_manual(values = c("#316593","#df7a26"), labels = c("outpatient", "inpatient"))

ST_bar_plot_inpatients_all_st

patient_and_mlst$grouped_st <- factor(patient_and_mlst$grouped_st, 
                              levels = names(sort(table(patient_and_mlst$grouped_st), 
                              decreasing=TRUE)))

ST_bar_plot_inpatients <- ggplot(filter(patient_and_mlst, !is.na(inpatient)), 
       aes(x = grouped_st, fill = factor(inpatient, levels = c(0,1)))) + 
       geom_bar(stat = "count") +
       theme(axis.text.x = element_text(angle = 90)) + 
       ggtitle("STs collected by inpatient/outpatient") +
       coord_flip() + 
       theme_light() + 
       ylab("Isolate count") +
       xlab("Sequence type") +
       theme(text = element_text(size = 22),
             legend.title = element_blank(),
             legend.position = c(0.5,0.5),
             legend.key.size = unit(20, "pt")) + 
       scale_fill_manual(values = c("#6aa2c6","#df7a26"), labels = c("outpatient", "inpatient"))

ST_bar_plot_inpatients
```

And then hospital-onset vs. not

```{r plot-st-by-hospital-onset, message = FALSE, warning = FALSE, fig.height = 6}
############################################################
# 1. ALL STs (sorted_ST) – community- vs healthcare-onset
############################################################

# ensure the factor levels are ordered 0 = community, 1 = healthcare
patient_and_mlst_filtered <- patient_and_mlst_filtered %>% 
  mutate(hospital_onset = factor(hospital_onset, levels = c(0, 1)))

ST_bar_plot_onset_all_st <- ggplot(
  filter(patient_and_mlst_filtered, !is.na(hospital_onset)),
  aes(x = sorted_ST, fill = hospital_onset)
) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("STs collected by community- vs healthcare-onset") +
  coord_flip() +
  theme_light() +
  ylab("Isolate count") +
  xlab("Sequence type") +
  theme(
    text = element_text(size = 22),
    legend.title = element_blank(),
    legend.position = c(0.5, 0.5),
    legend.key.size = unit(20, "pt")
  ) +
  scale_fill_manual(
    values = c("#316593", "#FF842A"),
    labels = c("community-onset", "healthcare-onset")
  )

ST_bar_plot_onset_all_st

############################################################
# 2. COMMON STs (>50 isolates) – community- vs healthcare-onset
############################################################

# keep the grouped_st factor ordered by overall frequency
patient_and_mlst$grouped_st <- factor(
  patient_and_mlst$grouped_st,
  levels = names(sort(table(patient_and_mlst$grouped_st), decreasing = TRUE))
)

ST_bar_plot_onset_grouped <- ggplot(
  filter(patient_and_mlst, !is.na(hospital_onset)),
  aes(x = grouped_st,
      fill = factor(hospital_onset, levels = c(0, 1)))
) +
  geom_bar(stat = "count") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle("Common STs collected by community- vs healthcare-onset") +
  coord_flip() +
  theme_light() +
  ylab("Isolate count") +
  xlab("Sequence type") +
  theme(
    text = element_text(size = 22),
    legend.title = element_blank(),
    legend.position = c(0.5, 0.5),
    legend.key.size = unit(20, "pt")
  ) +
  scale_fill_manual(
    values = c("#538797", "#FF842A"),
    labels = c("community-onset", "healthcare-onset")
  )

ST_bar_plot_onset_grouped
```

Next I looked at days until infection onset by sequence type for those isolates that were collected from inpatients (note does not restrict to hospital-onset only)

```{r days-between-admit-onset-1, echo = FALSE}
ggplot(filter(patient_and_mlst, inpatient == 1),
       aes(x = grouped_st, y = days_btwn_admit_and_onset)) + 
       geom_point() + 
       geom_boxplot()
```

The raw data has outliers that make it hard to visualize. Removing those: 

```{r days-between-admit-onset-2, echo = FALSE}
ggplot(filter(patient_and_mlst, inpatient == 1 & days_btwn_admit_and_onset < 300),
       aes(x = grouped_st, y = days_btwn_admit_and_onset)) + 
       geom_point() + 
       geom_boxplot()
```

Still pretty scrunched at the bottom. Let's look at a table instead: 

```{r days-between-admit-onset-table, warning = FALSE, message = FALSE, echo = FALSE}
kable(patient_and_mlst %>%
  filter(inpatient == 1) %>%
  group_by(grouped_st) %>%
  summarize(median = quantile(days_btwn_admit_and_onset, .5),
            mean = mean(days_btwn_admit_and_onset),
            sd = sd(days_btwn_admit_and_onset)))
```


## plot timing and recurrence
```{r plot-timing-recurrence, message = FALSE, warning = FALSE, fig.height = 6}
timing <- ggplot(filter(patient_and_mlst, grouped_st != "other"), aes(x = collection_date, y = grouped_st)) +
  geom_point(size = 2, alpha = 0.8) +
  xlab("Collection date") +
  ylab("Sequence type") +
  theme_light() + 
  ggtitle("Collection timing of newly sequenced C. difficile isolates") + 
  theme(text = element_text(size = 22))

timing
```

## Summary Table
```{r}
library(table1)
library(dplyr)
library(labelled)

# Format categorical variable for clarity
combined_data_single_isolate$GenderCode <- factor(combined_data_single_isolate$GenderCode,
                                                  levels = c("F", "M"),
                                                  labels = c("Female", "Male"))

combined_data_single_isolate$inpatient <- factor(combined_data_single_isolate$inpatient,
                                                 levels = c(0, 1),
                                                 labels = c("Outpatient", "Inpatient"))

combined_data_single_isolate$hospital_onset <- factor(combined_data_single_isolate$hospital_onset,
                                                      levels = c(0, 1),
                                                      labels = c("Community", "Hospital"))

# Assign labels
label(combined_data_single_isolate$AdmitAge) <- "Age"
label(combined_data_single_isolate$BMI) <- "BMI"
label(combined_data_single_isolate$AdmissionType) <- "Admission Type"
label(combined_data_single_isolate$GenderCode) <- "Sex"
label(combined_data_single_isolate$RaceName) <- "Race"
label(combined_data_single_isolate$LOS) <- "Length of Stay (days)"
label(combined_data_single_isolate$inpatient) <- "Patient Status"
label(combined_data_single_isolate$hospital_onset) <- "Onset Type"

# Generate Table 1 (no grouping)
table1(~ AdmitAge + BMI + AdmissionType + 
         GenderCode + RaceName + LOS + inpatient + hospital_onset,
       data = combined_data_single_isolate)
```

# with cluster data
```{r}
library(readr)
library(dplyr)
library(here)

# --- Clade 2 ---
clade2_clusters_df <- read_csv(here("data", "metadata", "clade2_clusters_df.csv"), show_col_types = FALSE)
names(clade2_clusters_df)[1] <- "genomeID"

clade2_clusters_df <- clade2_clusters_df %>%
  mutate(clade = "II",
         clade_cluster = paste0(clade, "-", cluster))

# --- Clade 1 ---
clade1_clusters_df <- read_csv(here("data", "metadata", "clade1_clusters_df.csv"), show_col_types = FALSE)
names(clade1_clusters_df)[1] <- "genomeID"

clade1_clusters_df <- clade1_clusters_df %>%
  mutate(clade = "I",
         clade_cluster = paste0(clade, "-", cluster))

# --- Combine ---
combined_clusters_df <- bind_rows(clade1_clusters_df, clade2_clusters_df)

# --- Save ---
dir.create(here("data", "metadata"), showWarnings = FALSE, recursive = TRUE)
write_csv(combined_clusters_df, here("data", "metadata", "combined_clusters_df.csv"))
```
## combine with demographics information
```{r}
# Join with combined_data
combined_data_with_clusters <- combined_data %>%
  left_join(combined_clusters_df, by = "genomeID")

# Join with combined_data_single_isolate (isolate == "a")
combined_data_single_isolate_with_clusters <- combined_data_single_isolate %>%
  left_join(combined_clusters_df, by = "genomeID")
```

## Summary Table
```{r}
library(table1)
library(dplyr)
library(labelled)

# Format categorical variable for clarity
combined_data_single_isolate_with_clusters$GenderCode <- factor(combined_data_single_isolate_with_clusters$GenderCode,
                                                  levels = c("F", "M"),
                                                  labels = c("Female", "Male"))

combined_data_single_isolate_with_clusters$inpatient <- factor(combined_data_single_isolate_with_clusters$inpatient,
                                                 levels = c(0, 1),
                                                 labels = c("Outpatient", "Inpatient"))

combined_data_single_isolate_with_clusters$hospital_onset <- factor(combined_data_single_isolate_with_clusters$hospital_onset,
                                                      levels = c(0, 1),
                                                      labels = c("Community", "Hospital"))

# Assign labels
label(combined_data_single_isolate_with_clusters$AdmitAge) <- "Age"
label(combined_data_single_isolate_with_clusters$BMI) <- "BMI"
label(combined_data_single_isolate_with_clusters$AdmissionType) <- "Admission Type"
label(combined_data_single_isolate_with_clusters$GenderCode) <- "Sex"
label(combined_data_single_isolate_with_clusters$RaceName) <- "Race"
label(combined_data_single_isolate_with_clusters$LOS) <- "Length of Stay (days)"
label(combined_data_single_isolate_with_clusters$inpatient) <- "Patient Status"
label(combined_data_single_isolate_with_clusters$hospital_onset) <- "Onset Type"

# Generate Table 1 (no grouping)
table1(~ AdmitAge + BMI + AdmissionType + 
         GenderCode + RaceName + LOS + inpatient + hospital_onset,
       data = combined_data_single_isolate_with_clusters %>% filter(!is.na(clade)))
```


# Initial exploration of locations and linkages
```{r, message=FALSE, warning=FALSE, results='hide'}
library("tidyverse")
library("lubridate")
library("here")
library("data.table")
library("ggthemes")
library("ggpubr")
library("openxlsx")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# patient locations
locations <- read_tsv(here('data', 'metadata', 'locations_from_wiens_group_filtered.tsv'))

#patient ids
pateint_ids <- read_csv(here('data', 'metadata', 'sampleid_patientid.csv'))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
ids_filtered <- pateint_ids %>%
  filter(sample_id %in% patient_data$sample_id)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
locations_filtered <- locations %>% 
  left_join(patient_data, by = c("PatientID" = "RDW_patient_id")) %>% 
  filter(EndDate > StartDate, StartDate >= "2016-02-02" & StartDate <= "2017-12-31")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
# Filter to rows where it seems like C diff was collected during that location row 
locations_where_cdiff_collected <- locations_filtered %>%
  filter(as.Date(collection_date) == as.Date(Date)) %>%
  filter(collection_date > StartDate & collection_date < EndDate) %>%
  distinct()
```

```{r, message=FALSE, warning=FALSE, results='hide'}
locations_with_st <- locations_filtered %>% 
  left_join(patient_and_mlst, by = c("PatientID" = "RDW_patient_id")) %>% 
  filter(!is.na(ST), EndDate > StartDate, StartDate < "2018-01-01") %>%
  select(-ends_with(".y")) %>%
  rename_with(~ sub("\\.x$", "", .), ends_with(".x")) %>%
  distinct()
```

## Locations distribution
```{r}
library(dplyr)

# Total unique patients
total_patients <- locations_filtered %>%
  summarise(unique_patients = n_distinct(PatientID))

# How many patients were in each Unit
unit_patient_counts <- locations_filtered %>%
  select(PatientID, Unit) %>%
  distinct() %>%
  count(Unit, name = "n_patients") %>%
  arrange(desc(n_patients))

# View results
total_patients
unit_patient_counts
```
## Overall Overlaps
```{r, message=FALSE, warning=FALSE, results='hide'}
library(data.table)

# Convert to data.table
df <- locations_with_st %>%
  select(PatientID, FacilityCode, Unit, Room, StartDate, EndDate, ST, clade, sorted_ST, secondary_isolate) %>%
  distinct()
dt <- as.data.table(df)
setkey(dt, StartDate, EndDate)

# Self join to find overlaps
overlaps <- foverlaps(dt, dt, by.x = c("StartDate", "EndDate"), type = "any", nomatch = 0L)

# Remove self-matches
overlaps <- overlaps[PatientID != i.PatientID]

# Create unique pair IDs
overlaps[, pair_id := paste0(pmin(PatientID, i.PatientID), "_", pmax(PatientID, i.PatientID))]
overlaps <- unique(overlaps, by = "pair_id")
```

```{r, message=FALSE, warning=FALSE, results='hide'}
facility_overlap <- overlaps[FacilityCode == i.FacilityCode]

unit_overlap <- overlaps[FacilityCode == i.FacilityCode & Unit == i.Unit]

room_overlap <- overlaps[FacilityCode == i.FacilityCode & Unit == i.Unit & Room == i.Room]
```

```{r}
library(data.table)
library(dplyr)

# overlaps: full overlap table from foverlaps
# facility_overlap, unit_overlap, room_overlap: filtered subsets

facility_summary <- facility_overlap[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode)
]
facility_summary

unit_summary <- unit_overlap[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode, Unit)
]
unit_summary

room_summary <- room_overlap[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode, Unit, Room)
]
room_summary
```
## Overlap with same ST

```{r}
library(data.table)
library(dplyr)

# 1. Filter for overlapping pairs where STs match
facility_overlap_sameST <- overlaps[
  FacilityCode == i.FacilityCode & ST == i.ST
]

unit_overlap_sameST <- overlaps[
  FacilityCode == i.FacilityCode & Unit == i.Unit & ST == i.ST
]

room_overlap_sameST <- overlaps[
  FacilityCode == i.FacilityCode & Unit == i.Unit & Room == i.Room & ST == i.ST
]

# 2. Summarize overlapping pairs with same ST at each level
facility_summary_sameST <- facility_overlap_sameST[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode)
]
facility_summary_sameST

unit_summary_sameST <- unit_overlap_sameST[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode, Unit)
]
unit_summary_sameST

room_summary_sameST <- room_overlap_sameST[
  , .(n_pairs = uniqueN(pair_id)), 
  by = .(FacilityCode, Unit, Room)
]
room_summary_sameST 
```
## Sequential Overlap Detection
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library("readr")
library("tidyverse")
library("readxl")
library("here")
library("RColorBrewer")
library("knitr")
library("ggthemes")
library(data.table)
```

```{r import-data, warning = FALSE, message = FALSE}
############################
# import relevant datasets
############################
ids <- read_csv(here('data','metadata','cleaned_passed_genomeid.csv'), col_names = FALSE)$X1

patient_data <- read_csv(here("data","metadata", "hospital_onset_data.csv")) %>%
  filter(genome_id %in% ids)

mlst_data <- read_tsv(here("data", "metadata", "mlst_replaced.tsv")) %>%
  mutate(ST = str_remove(ST, "\\*"))

main_wiens<- read_tsv(here("data", "metadata", "main_from_wiens_group_filtered_cdiff_cases.tsv")) %>%
  arrange(PatientID, Date)

# patient locations
locations <- read_tsv(here('data', 'metadata', 'locations_from_wiens_group_filtered.tsv'))

#patient ids
pateint_ids <- read_csv(here('data', 'metadata', 'sampleid_patientid.csv'))
```

```{r, message=FALSE, warning=FALSE, results='hide'}
ids_filtered <- pateint_ids %>%
  filter(sample_id %in% patient_data$sample_id)
```

```{r, message=FALSE, warning=FALSE, results='hide'}
locations_filtered <- locations %>% 
  left_join(patient_data, by = c("PatientID" = "RDW_patient_id")) %>% 
  filter(EndDate > StartDate, StartDate >= "2016-02-02" & StartDate <= "2017-12-31")
```

```{r merge-datasets, warning = FALSE, message = FALSE, echo = FALSE}
##############################
# to those that are sequenced
##############################
patient_and_mlst <- inner_join(patient_data, mlst_data, by = "genome_id")
```

```{r}
locations_with_st <- locations_filtered %>% 
  left_join(patient_and_mlst, by = c("PatientID" = "RDW_patient_id")) %>% 
  filter(!is.na(ST), EndDate > StartDate, StartDate < "2018-01-01") %>%
  select(-ends_with(".y")) %>%
  rename_with(~ sub("\\.x$", "", .), ends_with(".x")) %>%
  distinct()
# Convert to data.table
df <- locations_with_st %>%
  select(PatientID, FacilityCode, Unit, Room, StartDate, EndDate, ST, clade, sorted_ST, secondary_isolate) %>%
  distinct()
dt <- as.data.table(df)
```

### Same Unit
```{r}
# Sort by Facility, Unit, StartDate
setorder(dt, FacilityCode, Unit, StartDate)

# Within each unit, find next stay (ignore Room)
sequential_unit <- dt[, .SD[order(StartDate)], by = .(FacilityCode, Unit)][
  , `:=`(next_PatientID = shift(PatientID, type = "lead"),
         next_StartDate = shift(StartDate, type = "lead"),
         next_EndDate = shift(EndDate, type = "lead"),
         next_ST = shift(ST, type = "lead"),
         next_clade = shift(clade, type = "lead"),
         next_Room = shift(Room, type = "lead")),
  by = .(FacilityCode, Unit)
][
  !is.na(next_PatientID) & PatientID != next_PatientID & EndDate <= next_StartDate
][
  , time_between_days := as.integer(difftime(next_StartDate, EndDate, units = "days"))
]

# Show sequential transitions with ≤30 days between stays
sequential_transfers_unit <- sequential_unit[time_between_days <= 30]

# Example output
sequential_transfers_unit[, .(FacilityCode, Unit, Room,
                         from = PatientID, to = next_PatientID,
                         from_end = EndDate, to_start = next_StartDate,
                         time_between_days,
                         ST, next_ST, clade, next_clade)]
```

```{r}
sameST_unit <- sequential_transfers_unit[ST == next_ST]
sameST_unit[, .(FacilityCode, Unit, Room,
                         from = PatientID, to = next_PatientID,
                         from_end = EndDate, to_start = next_StartDate,
                         time_between_days,
                         ST, next_ST, clade, next_clade)]
```

### Same Room
```{r}
# Sort by Facility, Unit, Room, StartDate
setorder(dt, FacilityCode, Unit, Room, StartDate)

# Within each room, find next stay
sequential_room <- dt[, .SD[order(StartDate)], by = .(FacilityCode, Unit, Room)][
  , `:=`(next_PatientID = shift(PatientID, type = "lead"),
         next_StartDate = shift(StartDate, type = "lead"),
         next_EndDate = shift(EndDate, type = "lead"),
         next_ST = shift(ST, type = "lead"),
         next_clade = shift(clade, type = "lead")),
  by = .(FacilityCode, Unit, Room)
][
  !is.na(next_PatientID) & PatientID != next_PatientID & EndDate <= next_StartDate
][
  , time_between_days := as.integer(difftime(next_StartDate, EndDate, units = "days"))
]

# Show sequential transitions with ≤30 days between stays
sequential_transfers_room <- sequential_room[time_between_days <= 30]

# Example output
sequential_transfers_room[, .(FacilityCode, Unit, Room,
                         from = PatientID, to = next_PatientID,
                         from_end = EndDate, to_start = next_StartDate,
                         time_between_days,
                         ST, next_ST, clade, next_clade)]
```

```{r}
sameST_room <- sequential_transfers_room[ST == next_ST]
sameST_room[, .(FacilityCode, Unit, Room,
                         from = PatientID, to = next_PatientID,
                         from_end = EndDate, to_start = next_StartDate,
                         time_between_days,
                         ST, next_ST, clade, next_clade)]
```

### Visualization
```{r}
# Long format for Unit
unit_long <- rbind(
  sequential_unit[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sequential_unit[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                      StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

unit_long[, UnitID := paste(FacilityCode, Unit, sep = "_")]


## Long format for Room
room_long <- rbind(
  sequential_room[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sequential_room[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                      StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

room_long[, RoomID := paste(FacilityCode, Unit, Room, sep = "_")]
```
### Sequential overlaps by Unit
```{r}
ggplot(unit_long, aes(x = StartDate, xend = EndDate, 
                      y = UnitID, yend = UnitID, 
                      color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```

### Sequential overlaps by Room
```{r}
ggplot(unit_long, aes(x = StartDate, xend = EndDate, 
                      y = UnitID, yend = UnitID, 
                      color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```
### Same ST by unit
```{r}
unit_long_sameST <- rbind(
  sameST_unit[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sameST_unit[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                  StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

unit_long_sameST[, UnitID := paste(FacilityCode, Unit, sep = "_")]

ggplot(unit_long_sameST, aes(x = StartDate, xend = EndDate, 
                             y = UnitID, yend = UnitID, 
                             color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit (Same ST)", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```
### Same ST by Room
```{r}
room_long_sameST <- rbind(
  sameST_room[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sameST_room[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                  StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

room_long_sameST[, RoomID := paste(FacilityCode, Unit, Room, sep = "_")]

ggplot(room_long_sameST, aes(x = StartDate, xend = EndDate, 
                             y = RoomID, yend = RoomID, 
                             color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Room (Same ST)", 
       x = "Date", y = "Room", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))
```
### Summary Table
```{r}
library(data.table)

# 1. Summary for sequential overlaps in same Unit
summary_unit <- sequential_unit[
  , .(
    n_pairs = .N,
    min_days = min(time_between_days, na.rm = TRUE),
    max_days = max(time_between_days, na.rm = TRUE),
    mean_days = round(mean(time_between_days, na.rm = TRUE), 1),
    median_days = median(time_between_days, na.rm = TRUE)
  )
]

# 2. Summary for sequential overlaps in same Room
summary_room <- sequential_room[
  , .(
    n_pairs = .N,
    min_days = min(time_between_days, na.rm = TRUE),
    max_days = max(time_between_days, na.rm = TRUE),
    mean_days = round(mean(time_between_days, na.rm = TRUE), 1),
    median_days = median(time_between_days, na.rm = TRUE)
  )
]

# 3. Combine into one summary table
summary_table <- rbind(
  data.table(type = "Same Unit", summary_unit),
  data.table(type = "Same Room", summary_room)
)

# 4. Display
summary_table
```

### Sensitivity Test
```{r}
cutoffs <- c(7, 14, 30, 60)

unit_sensitivity <- lapply(cutoffs, function(cutoff) {
  sequential_unit[time_between_days <= cutoff, .N]
}) %>%
  unlist()

unit_sensitivity_df <- data.table(
  Cutoff_Days = cutoffs,
  N_Pairs = unit_sensitivity,
  Total_Pairs = nrow(sequential_unit),
  Percent_Included = round(100 * unit_sensitivity / nrow(sequential_unit), 1)
)

room_sensitivity <- lapply(cutoffs, function(cutoff) {
  sequential_room[time_between_days <= cutoff, .N]
}) %>%
  unlist()

room_sensitivity_df <- data.table(
  Cutoff_Days = cutoffs,
  N_Pairs = room_sensitivity,
  Total_Pairs = nrow(sequential_room),
  Percent_Included = round(100 * room_sensitivity / nrow(sequential_room), 1)
)

unit_sensitivity_df[, Type := "Same Unit"]
room_sensitivity_df[, Type := "Same Room"]

sensitivity_summary <- rbind(unit_sensitivity_df, room_sensitivity_df)
sensitivity_summary
```
```{r}
library(ggplot2)

ggplot(sensitivity_summary, aes(x = Cutoff_Days, y = Percent_Included, color = Type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Sensitivity Analysis of Sequential Overlap Cutoffs",
       x = "Cutoff (days between stays)",
       y = "Percent of Sequential Pairs Included",
       color = "Type") +
  scale_x_continuous(breaks = cutoffs) +
  theme_minimal()
```
#### Same ST
```{r}
summary_sameST_unit <- sameST_unit[
  , .(
    n_pairs = .N,
    min_days = min(time_between_days, na.rm = TRUE),
    max_days = max(time_between_days, na.rm = TRUE),
    mean_days = round(mean(time_between_days, na.rm = TRUE), 1),
    median_days = median(time_between_days, na.rm = TRUE)
  )
]

summary_sameST_room <- sameST_room[
  , .(
    n_pairs = .N,
    min_days = min(time_between_days, na.rm = TRUE),
    max_days = max(time_between_days, na.rm = TRUE),
    mean_days = round(mean(time_between_days, na.rm = TRUE), 1),
    median_days = median(time_between_days, na.rm = TRUE)
  )
]

summary_sameST_table <- rbind(
  data.table(type = "Same Unit (Same ST)", summary_sameST_unit),
  data.table(type = "Same Room (Same ST)", summary_sameST_room)
)

summary_sameST_table
```
```{r}
cutoffs <- c(7, 14, 30)

sameST_unit_sensitivity <- lapply(cutoffs, function(cutoff) {
  sameST_unit[time_between_days <= cutoff, .N]
}) %>% unlist()

sameST_unit_df <- data.table(
  Cutoff_Days = cutoffs,
  N_Pairs = sameST_unit_sensitivity,
  Total_Pairs = nrow(sameST_unit),
  Percent_Included = round(100 * sameST_unit_sensitivity / nrow(sameST_unit), 1),
  Type = "Same Unit (Same ST)"
)

sameST_room_sensitivity <- lapply(cutoffs, function(cutoff) {
  sameST_room[time_between_days <= cutoff, .N]
}) %>% unlist()

sameST_room_df <- data.table(
  Cutoff_Days = cutoffs,
  N_Pairs = sameST_room_sensitivity,
  Total_Pairs = nrow(sameST_room),
  Percent_Included = round(100 * sameST_room_sensitivity / nrow(sameST_room), 1),
  Type = "Same Room (Same ST)"
)

sameST_sensitivity_table <- rbind(sameST_unit_df, sameST_room_df)
sameST_sensitivity_table
```

### Visualization (30 days cutoff)
### Sequential overlaps by Unit
```{r}
ggplot(unit_long, aes(x = StartDate, xend = EndDate, 
                      y = UnitID, yend = UnitID, 
                      color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```

### Sequential overlaps by Room
```{r}
ggplot(unit_long, aes(x = StartDate, xend = EndDate, 
                      y = UnitID, yend = UnitID, 
                      color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```
### Same ST by unit
```{r}
unit_long_sameST <- rbind(
  sameST_unit[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sameST_unit[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                  StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

unit_long_sameST[, UnitID := paste(FacilityCode, Unit, sep = "_")]

ggplot(unit_long_sameST, aes(x = StartDate, xend = EndDate, 
                             y = UnitID, yend = UnitID, 
                             color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Unit (Same ST)", 
       x = "Date", y = "Unit", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```
### Same ST by Room
```{r}
room_long_sameST <- rbind(
  sameST_room[, .(FacilityCode, Unit, Room, PatientID, StartDate, EndDate, ST)],
  sameST_room[, .(FacilityCode, Unit, Room, PatientID = next_PatientID,
                  StartDate = next_StartDate, EndDate = next_EndDate, ST = next_ST)]
) %>% distinct()

room_long_sameST[, RoomID := paste(FacilityCode, Unit, Room, sep = "_")]

ggplot(room_long_sameST, aes(x = StartDate, xend = EndDate, 
                             y = RoomID, yend = RoomID, 
                             color = as.factor(ST))) +
  geom_segment(size = 3) +
  labs(title = "Sequential Stays in Same Room (Same ST)", 
       x = "Date", y = "Room", color = "ST") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 5))
```
