---
title: "Build Pair Table for Case-Control Analysis (Clade 2)"
author: "Skylar Zeng"
date: "`r format(Sys.Date())`"
output: html_document
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(lubridate)
library(here)
library(readr)
library(janitor)
```

## Paths and Parameters
```{r}
# Master linkage table
master_linkage <- here("2025_analysis", "isolate_master.csv")

# SNV distance files (clade 2)
post_noncore <- here("data", "SNV_matrices", "clade2_post_noncore_snv_long.csv")
post_softcore <- here("data", "SNV_matrices", "clade2_post_softcore_snv_long.csv")
pre_core <- here("data", "SNV_matrices", "clade2_pre_core.csv")

# MLST file
mlst_file <- here("data", "metadata", "mlst_replaced.tsv")

# Output directory
output_dir <- here("2025_analysis")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
output_file <- here(output_dir, "pairs_core.csv")

# Parameters
snv_case_threshold <- 2      # Cases: SNV ≤2
snv_control_threshold_1 <- 10 # Controls: SNV >10
snv_control_threshold_2 <- 20 # Controls: SNV >20 (for sensitivity analysis)
max_controls_per_case <- NULL # Set to a number (e.g., 10) to cap controls, or NULL for no cap
set.seed(42)  # For reproducible random sampling if capping controls
```

# Load Data

## Load master linkage table and filter to clade 2
```{r}
master <- read_csv(master_linkage, show_col_types = FALSE) %>%
  filter(clade == 2) %>%
  filter(!is.na(st))  # Only genomes with ST assigned

cat("Clade 2 genomes with ST assigned:", nrow(master), "\n")
cat("Unique STs in clade 2:", n_distinct(master$st), "\n")

# Get list of valid genome IDs for clade 2
clade2_genome_ids <- master$genome_id
cat("Valid clade 2 genome IDs:", length(clade2_genome_ids), "\n")
```

## Load SNV distance files
```{r}
# Function to normalize genome IDs (handle CDIF_formed_ vs CDIF_SB_)
normalize_genome_id <- function(id) {
  # Remove reference genome entries
  is_ref <- grepl("ref_genome|complete genome|R20291", id, ignore.case = TRUE)
  # Return NA for reference genomes, otherwise keep ID as-is
  ifelse(is_ref, NA_character_, id)
}

# Pre-core (wide format with reference genome)
pre_core_raw <- read_csv(pre_core, show_col_types = FALSE) %>%
  rename(snv_pre = snvs)

# Post-noncore (long format)
post_noncore_raw <- read_csv(post_noncore, show_col_types = FALSE) %>%
  rename(snv_post_noncore = snvs)

# Post-softcore (long format)
post_softcore_raw <- read_csv(post_softcore, show_col_types = FALSE) %>%
  rename(snv_post_softcore = snvs)

cat("Pre-core pairs:", nrow(pre_core_raw), "\n")
cat("Post-noncore pairs:", nrow(post_noncore_raw), "\n")
cat("Post-softcore pairs:", nrow(post_softcore_raw), "\n")
```

## Process SNV distance data
```{r}
# Get all valid genome IDs (both CDIF_SB_ and CDIF_formed_ from master, plus check SNV files)
all_snv_ids <- c(
  pre_core_raw$id1, pre_core_raw$id2,
  post_noncore_raw$id1, post_noncore_raw$id2,
  post_softcore_raw$id1, post_softcore_raw$id2
) %>%
  unique() %>%
  .[!grepl("ref_genome|complete genome|R20291", ., ignore.case = TRUE)]

# Combine with master table IDs
clade2_genome_ids_all <- unique(c(clade2_genome_ids, all_snv_ids))
cat("Total unique genome IDs to consider:", length(clade2_genome_ids_all), "\n")

# Process pre-core: filter to isolate pairs only (exclude reference genome)
pre_core_clean <- pre_core_raw %>%
  mutate(
    id1_clean = normalize_genome_id(id1),
    id2_clean = normalize_genome_id(id2)
  ) %>%
  filter(!is.na(id1_clean) & !is.na(id2_clean)) %>%
  filter(id1_clean %in% clade2_genome_ids_all & id2_clean %in% clade2_genome_ids_all) %>%
  select(id1 = id1_clean, id2 = id2_clean, snv_pre) %>%
  # Ensure id1 < id2 to avoid duplicates
  mutate(
    id1_new = pmin(id1, id2),
    id2_new = pmax(id1, id2)
  ) %>%
  select(id1 = id1_new, id2 = id2_new, snv_pre) %>%
  distinct()

# Process post-noncore
post_noncore_clean <- post_noncore_raw %>%
  mutate(
    id1_clean = normalize_genome_id(id1),
    id2_clean = normalize_genome_id(id2)
  ) %>%
  filter(!is.na(id1_clean) & !is.na(id2_clean)) %>%
  filter(id1_clean %in% clade2_genome_ids_all & id2_clean %in% clade2_genome_ids_all) %>%
  select(id1 = id1_clean, id2 = id2_clean, snv_post_noncore) %>%
  mutate(
    id1_new = pmin(id1, id2),
    id2_new = pmax(id1, id2)
  ) %>%
  select(id1 = id1_new, id2 = id2_new, snv_post_noncore) %>%
  distinct()

# Process post-softcore
post_softcore_clean <- post_softcore_raw %>%
  mutate(
    id1_clean = normalize_genome_id(id1),
    id2_clean = normalize_genome_id(id2)
  ) %>%
  filter(!is.na(id1_clean) & !is.na(id2_clean)) %>%
  filter(id1_clean %in% clade2_genome_ids_all & id2_clean %in% clade2_genome_ids_all) %>%
  select(id1 = id1_clean, id2 = id2_clean, snv_post_softcore) %>%
  mutate(
    id1_new = pmin(id1, id2),
    id2_new = pmax(id1, id2)
  ) %>%
  select(id1 = id1_new, id2 = id2_new, snv_post_softcore) %>%
  distinct()

cat("\nAfter filtering to clade 2 isolates:\n")
cat("Pre-core pairs:", nrow(pre_core_clean), "\n")
cat("Post-noncore pairs:", nrow(post_noncore_clean), "\n")
cat("Post-softcore pairs:", nrow(post_softcore_clean), "\n")
```

## Join SNV distances and add ST information
```{r}
# Load MLST data to get ST for all genomes (including CDIF_formed_)
mlst_all <- read_tsv(mlst_file, show_col_types = FALSE) %>%
  clean_names() %>%
  select(genome_id, st, clade) %>%
  filter(clade == 2) %>%
  filter(!is.na(st))

cat("MLST data: clade 2 genomes with ST:", nrow(mlst_all), "\n")

# Combine all SNV distances
snv_all <- pre_core_clean %>%
  full_join(post_noncore_clean, by = c("id1", "id2")) %>%
  full_join(post_softcore_clean, by = c("id1", "id2"))

# Add ST information for both genomes
mlst_st <- mlst_all %>%
  select(genome_id, st) %>%
  rename(st1 = st)

snv_with_st <- snv_all %>%
  left_join(mlst_st, by = c("id1" = "genome_id")) %>%
  left_join(mlst_st %>% rename(st2 = st1), by = c("id2" = "genome_id")) %>%
  # Filter to within-ST pairs only
  filter(!is.na(st1) & !is.na(st2)) %>%
  filter(st1 == st2) %>%
  mutate(st = st1) %>%
  select(id1, id2, st, snv_pre, snv_post_noncore, snv_post_softcore)

cat("\nWithin-ST pairs (clade 2):\n")
cat("Total pairs:", nrow(snv_with_st), "\n")
cat("STs represented:", n_distinct(snv_with_st$st), "\n")
```

# Build Pair Table

## Create pair table for each pipeline
```{r}
# Function to create pair table for a given pipeline
create_pair_table <- function(snv_col, pipeline_name) {
  pairs <- snv_with_st %>%
    mutate(snv = .data[[snv_col]]) %>%
    select(id1, id2, st, snv) %>%
    filter(!is.na(snv)) %>%
    # Classify as case or control
    mutate(
      case_control = case_when(
        snv <= snv_case_threshold ~ "case",
        snv > snv_control_threshold_1 ~ "control_gt10",
        TRUE ~ NA_character_  # Exclude pairs in between (3-10)
      )
    ) %>%
    filter(!is.na(case_control)) %>%
    # Add control_gt20 flag for sensitivity analysis
    mutate(
      control_gt20 = if_else(snv > snv_control_threshold_2, TRUE, FALSE)
    ) %>%
    mutate(pipeline = pipeline_name)
  
  return(pairs)
}

# Create pair tables for each pipeline
pairs_pre <- create_pair_table("snv_pre", "pre_core")
pairs_post_noncore <- create_pair_table("snv_post_noncore", "post_noncore")
pairs_post_softcore <- create_pair_table("snv_post_softcore", "post_softcore")

cat("Pre-core pairs:\n")
cat("  Cases (SNV ≤2):", sum(pairs_pre$case_control == "case"), "\n")
cat("  Controls (SNV >10):", sum(pairs_pre$case_control == "control_gt10"), "\n")
cat("  Controls (SNV >20):", sum(pairs_pre$control_gt20), "\n")

cat("\nPost-noncore pairs:\n")
cat("  Cases (SNV ≤2):", sum(pairs_post_noncore$case_control == "case"), "\n")
cat("  Controls (SNV >10):", sum(pairs_post_noncore$case_control == "control_gt10"), "\n")
cat("  Controls (SNV >20):", sum(pairs_post_noncore$control_gt20), "\n")

cat("\nPost-softcore pairs:\n")
cat("  Cases (SNV ≤2):", sum(pairs_post_softcore$case_control == "case"), "\n")
cat("  Controls (SNV >10):", sum(pairs_post_softcore$case_control == "control_gt10"), "\n")
cat("  Controls (SNV >20):", sum(pairs_post_softcore$control_gt20), "\n")
```

## Optionally cap controls per case
```{r}
# If max_controls_per_case is set, randomly sample controls
cap_controls <- function(pairs_df) {
  if(is.null(max_controls_per_case)) {
    return(pairs_df)
  }
  
  # For each case, sample controls within the same ST
  cases <- pairs_df %>%
    filter(case_control == "case") %>%
    select(id1, id2, st, pipeline)
  
  controls <- pairs_df %>%
    filter(case_control == "control_gt10")
  
  # Sample controls per case
  controls_sampled <- cases %>%
    left_join(controls, by = c("st", "pipeline"), suffix = c("_case", "_control")) %>%
    group_by(id1_case, id2_case, st, pipeline) %>%
    sample_n(min(max_controls_per_case, n()), replace = FALSE) %>%
    ungroup() %>%
    select(id1 = id1_control, id2 = id2_control, st, snv, case_control, control_gt20, pipeline)
  
  # Combine cases and sampled controls
  pairs_capped <- bind_rows(
    pairs_df %>% filter(case_control == "case"),
    controls_sampled
  )
  
  return(pairs_capped)
}

# Apply capping if specified
if(!is.null(max_controls_per_case)) {
  cat("\nCapping controls to", max_controls_per_case, "per case...\n")
  pairs_pre <- cap_controls(pairs_pre)
  pairs_post_noncore <- cap_controls(pairs_post_noncore)
  pairs_post_softcore <- cap_controls(pairs_post_softcore)
  
  cat("After capping:\n")
  cat("Pre-core controls:", sum(pairs_pre$case_control == "control_gt10"), "\n")
  cat("Post-noncore controls:", sum(pairs_post_noncore$case_control == "control_gt10"), "\n")
  cat("Post-softcore controls:", sum(pairs_post_softcore$case_control == "control_gt10"), "\n")
}
```

## Combine all pipelines and add metadata
```{r}
# Combine all pair tables
pairs_all <- bind_rows(pairs_pre, pairs_post_noncore, pairs_post_softcore)

# Prepare master table for joining
master_join <- master %>%
  select(genome_id, patient_id, collection_date, st) %>%
  rename(
    patient_i = patient_id,
    collection_date_i = collection_date
  )

# Join metadata for genome_i (id1)
pairs_final <- pairs_all %>%
  left_join(master_join, by = c("id1" = "genome_id")) %>%
  rename(
    genome_i = id1,
    genome_j = id2,
    patient_i = patient_i,
    collection_date_i = collection_date_i,
    st_i = st.x
  )

# Join metadata for genome_j (id2)
master_join_j <- master %>%
  select(genome_id, patient_id, collection_date, st) %>%
  rename(
    patient_j = patient_id,
    collection_date_j = collection_date
  )

pairs_final <- pairs_final %>%
  left_join(master_join_j, by = c("genome_j" = "genome_id")) %>%
  # Use ST from the pair (should be same for both)
  mutate(st = st_i) %>%
  select(genome_i, genome_j, patient_i, patient_j, st, pipeline, snv, 
         case_control, collection_date_i, collection_date_j, control_gt20)

cat("\nFinal pair table:\n")
cat("Total pairs:", nrow(pairs_final), "\n")
cat("Cases:", sum(pairs_final$case_control == "case"), "\n")
cat("Controls (SNV >10):", sum(pairs_final$case_control == "control_gt10"), "\n")
cat("Controls (SNV >20):", sum(pairs_final$control_gt20, na.rm = TRUE), "\n")
cat("Pipelines:", paste(unique(pairs_final$pipeline), collapse = ", "), "\n")
```

## Summary statistics
```{r}
cat("\n=== PAIR TABLE SUMMARY ===\n\n")

# Summary by pipeline and case_control
summary_by_pipeline <- pairs_final %>%
  group_by(pipeline, case_control) %>%
  summarise(
    n_pairs = n(),
    mean_snv = round(mean(snv, na.rm = TRUE), 2),
    median_snv = median(snv, na.rm = TRUE),
    min_snv = min(snv, na.rm = TRUE),
    max_snv = max(snv, na.rm = TRUE),
    .groups = "drop"
  )

print(summary_by_pipeline)

# Summary by ST
summary_by_st <- pairs_final %>%
  filter(case_control == "case") %>%
  group_by(st, pipeline) %>%
  summarise(n_cases = n(), .groups = "drop") %>%
  group_by(st) %>%
  summarise(
    total_cases = sum(n_cases),
    pipelines = paste(unique(pipeline), collapse = ", "),
    .groups = "drop"
  ) %>%
  arrange(desc(total_cases))

cat("\nTop STs by number of case pairs:\n")
print(head(summary_by_st, 10))
```

# Save Output

## Full pair table (includes pairs with NA patient_id)
```{r}
# Select final columns as specified
pairs_output <- pairs_final %>%
  select(
    genome_i,
    genome_j,
    patient_i,
    patient_j,
    st,
    pipeline,
    snv,
    case_control,
    collection_date_i,
    collection_date_j
  ) %>%
  arrange(pipeline, st, case_control, snv)

# Save to CSV
write_csv(pairs_output, output_file)
cat("\n=== OUTPUT SAVED ===\n")
cat("Pair table saved to:", output_file, "\n")
cat("Total rows:", nrow(pairs_output), "\n")
cat("Pairs with NA patient_i:", sum(is.na(pairs_output$patient_i)), "\n")
cat("Pairs with NA patient_j:", sum(is.na(pairs_output$patient_j)), "\n")
cat("Pairs with NA in either patient_id:", sum(is.na(pairs_output$patient_i) | is.na(pairs_output$patient_j)), "\n")
cat("Columns:", paste(names(pairs_output), collapse = ", "), "\n")
```

## Filtered pair table (removes pairs with NA patient_id)
```{r}
# Filter out pairs where either patient_i or patient_j is NA
pairs_filtered <- pairs_output %>%
  filter(!is.na(patient_i) & !is.na(patient_j))

# Create output filename with _filtered suffix
output_file_filtered <- here(output_dir, "pairs_core_filtered.csv")

# Save filtered table
write_csv(pairs_filtered, output_file_filtered)

cat("\n=== FILTERED OUTPUT SAVED ===\n")
cat("Filtered pair table saved to:", output_file_filtered, "\n")
cat("Total rows (after filtering):", nrow(pairs_filtered), "\n")
cat("Rows removed:", nrow(pairs_output) - nrow(pairs_filtered), "\n")
cat("Percentage retained:", round(100 * nrow(pairs_filtered) / nrow(pairs_output), 2), "%\n")

# Summary of filtered data
cat("\nFiltered pair table summary:\n")
summary_filtered <- pairs_filtered %>%
  group_by(pipeline, case_control) %>%
  summarise(
    n_pairs = n(),
    mean_snv = round(mean(snv, na.rm = TRUE), 2),
    .groups = "drop"
  )

print(summary_filtered)
```

